Efficiently Handling Guards in the Low Level Design of RPython's tracing JIT

### 2012-08-01 talk/vmil2012/paper.tex
extend and refactor the introduction

%\begin{figure*}
%    \include{figures/ops_count_table}
%    \caption{Relative numbers of operations in the traces generated for
%    different benchmarks}
%    \label{fig:ops_count}
%\end{figure*}
Although there are several publications about tracing just-in-time compilers, to
our knowledge, there are none that describe the use and implementation of
guards in this context. With the following contributions we aim to shed some
light (to much?) on this topic.
The contributions of this paper are:
\todo{more motivation}
Based on the informal observation that guards are among the most common
operations in the traces produced by PyPy's tracing JIT and that guards are
operations that are associated with an overhead to maintain information about
state to be able to rebuild it, our goal is to present concrete numbers for the
frequency and the overhead produced by guards, explain how they are implemented
in the different levels of PyPy's tracing JIT and explain the rationale behind
the design decisions based on the numbers.
As can be seen on Figure~\ref{fig:ops_count} guards account for 14.42\% to
22.32\% of the operations before and for 15.2\% to 20.12\% of after the
optimization pass over the traced and compiled paths of the benchmarks.
Figure~\ref{fig:benchmarks} shows the absolute number of operations for each
benchmark, for every guard that stays alive after optimization there are
several kinds of metadata created and stored at different levels of the JIT to
be able to rebuild the interpreter or tracer state from a guard failure making
the optimization \bivab{some good word} of guards an important aspect of the
low-level design of a tracing just-in-time compiler.
The contributions of this paper are
In this paper we want to substantiate the aforementioned and describe based on
them the reasoning behind and the implementation of guards in PyPy's tracing
just-in-time compiler.
\begin{figure}
    \include{figures/guard_table}
    \caption{Percentage of guards before and after optimization for different benchmarks}
    \label{fig:guard_percent}
\end{figure}
\begin{itemize}
  \item An analysis of guards in the context of PyPy's tracing JIT to
  substantiate the aforementioned observation, based on a set of benchmarks.
  \item We provide a detailed measurements about the frequency and the
  overhead associated with guards.
  \item We provide a description about how guards are implemented in the high\-
  and low-level parts of the JIT and describe the rationale behind the design.
\end{itemize}

>>>>

%\begin{figure*}
%    \include{figures/ops_count_table}
%    \caption{Relative numbers of operations in the traces generated for
%    different benchmarks}
%    \label{fig:ops_count}
%\end{figure*}
Although there are several publications about tracing just-in-time compilers,
to our knowledge, there are none that describe deoptimization and the use and
implementation of guards in this context.

Based on the informal observation that guards are among the most common
operations in the traces produced by PyPy's tracing JIT and that guards are
operations that are associated with an overhead to maintain information about
state to be able to rebuild the execution state in case of deoptimization, our
goal is to present concrete numbers for the frequency and the overhead produced
by guards, explain how they are implemented in the different levels of PyPy's
tracing JIT and explain the rationale behind the design decisions based on the
numbers.

The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed, this process is described in more detail in
Section~\ref{sec:Resume Data}, during the recording phase special operations,
\texttt{guards}, are inserted into the recorded trace at all points where
control flow could diverge. As can be seen on Figure~\ref{fig:guard_percent}
guards account for 14.42\% to 22.32\% of the operations before and for 15.2\%
to 20.12\% of the operations after the optimization pass over the traced and
compiled parts of the benchmarks, making guards one of the most common
operations. Many of these guards fail rarely on not all during execution. Given
that associated with each guard information is stored, that is required to
rebuild the execution state in case control flow diverges from the recorded
path at a guard it is important to store the information associated with the
guards in a manner that tries to keep the overhead for storing the information
low while avoiding to put a burden on the execution of the recorded trace,
making the optimization of guards an important aspect of
the low-level design of a tracing just-in-time compiler.

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind and the implementation of guards in
PyPy's tracing just-in-time compiler, the contributions of this paper are:
\begin{itemize}
  \item An analysis of guards in the context of PyPy's tracing JIT to
  substantiate the aforementioned observation, based on a set of benchmarks.
  \item We provide a detailed measurements about the frequency and the
  overhead associated with guards.
  \item We provide a description about how guards are implemented in the high\-
  and low-level parts of the JIT and describe the rationale behind the design.
\end{itemize}
\begin{figure}
    \include{figures/guard_table}
    \caption{Percentage of guards before and after optimization for different benchmarks}
    \label{fig:guard_percent}
\end{figure}
